### 2017-10-13 energy_clustering/aistats18/energy_clustering.tex
energy to aistats

\documentclass[twoside]{article}
%\usepackage{aistats2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

@@

\aistatsauthor{Guilherme Fran\c{c}a \And  Joshua T. Vogelstein}

\aistatsaddress{ Johns Hopkins University } ]

\begin{abstract}
Energy statistics introduces the notion of potential
energy between probability distributions, in  
close analogy to Newton's gravitational
potential in physics. We propose a principled approach to the
clustering problem based on energy statistics theory.
Our mathematical formulation establishes connection to kernel methods,
leading to a quadratically constrained quadratic program (QCQP).
To obtain local solutions of such NP-hard QCQP 
we introduce an iterative
algorithm based
on Hartigan's method, which has the same computational cost but offers
several advantages compared to  
kernel $k$-means, based on Lloyd's heuristic.
We provide carefully designed numerical experiments showing
that our method is more flexible and outperforms
kernel $k$-means, spectral clustering,
standard $k$-means and Gaussian mixture models in a variety of settings,
specially in high dimensions. 
\end{abstract}

@@

Energy statistics \citep{Szkely2013}
provides a hypothesis test for equality of 
distributions which is achieved 
under minimum statistical potential energy. 
When probability distributions are different the 
statistical potential energy diverges as sample size increases, while tends 
to a nondegenerate limit distribution when probability
distributions are equal. 
The test statistic has compact representation
in terms of expectations of pairwise distances, providing
straightforward empirical estimates. Energy statistics
has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \citep{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \citep{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, an application of 
energy statistics to clustering, in Euclidean spaces, was 
proposed \citep{Kgroups}, which in part motivated this paper. 
We refer to \citep{Szkely2013} for an overview.

@@

Although clustering from energy statistics in Euclidean spaces was considered
in \citep{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. Our results
immediately establish the connection with kernel methods, showing that this 
QCQP is equivalent to kernel $k$-means optimization problem.

Our main 
algorithmic contribution is to use Hartigan's method \citep{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used in \citep{Kgroups}, without any
connection to kernels. More importantly, its 
advantages
over Lloyd's method was already demonstrated 
in some simple settings
\citep{Telgarsky,Slonin}, but apparently this method did not receive 
the deserved attention. To the 
best of our knowledge, Hartigan's method was not previously 
employed together with kernel methods. 
We provide a full kernel based Hartigan's algorithm for clustering,
where the kernel is fixed by energy statistics. 
We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to solve our QCQP. We show that both algorithms  have the same
time complexity, but Hartigan's method in kernel spaces offer several
advantages. Furthermore, in the examples considered in this paper, it 
also provides superior performance compared to a spectral clustering.
Our numerical results provide compelling evidence that 
Hartigan's method applied to energy statistics based clustering
is more accurate and robust than 
kernel $k$-means. Moreover, we illustrate the
flexibility of energy clustering,  
showing that it is able to perform accurately on data coming from 
different distributions, contrary to $k$-means and GMM for instance.
The proposed algorithm performs 
closely to $k$-means and GMM on normally distributed data, however,
it is significantly better on other settings. 
Its superiority in high dimensions is striking, being more accurate 
than $k$-means and GMM even on Gaussian settings.

@@

Consider random vectors $X_i$ living in an arbitrary space
$\mathcal{X}$ of \emph{negative type}, which means that $\mathcal{X}$
is endowed with a \emph{semimetric} 
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$ satisfying
$\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0$,
where $c_i \in \mathbb{R}$ and
$\sum_{i=1}^n c_i = 0$. 
Let $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments, and 
$X,X',Y,Y' \in \mathcal{X}$. 
The \emph{generalized energy distance} between $P$ and $Q$ is
given by 
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation} 
This quantity is nonnegative, $\Energy(P,Q) \ge 0$, and 
$\Energy^{1/2}$ is
a metric on the space of distributions. Energy distance provides
a characterization of equality of distributions.

For instance, the \emph{standard
energy distance} \citep{Szkely2013} in Euclidean spaces uses
the semimetric
\begin{equation}
\label{eq:rho_standard}
\rho_\alpha(X,Y) = \| X - Y\|^\alpha,
\end{equation} 
where $0< \alpha \le 2$ and $\| \cdot \|$ is the
Euclidean norm in $\mathcal{X}=\mathbb{R}^D$. In this case $\Energy(P,Q)$
is rotationally invariant. For $0<\alpha<2$ we have
$\Energy(P,Q) = 0$ if and only if $P=Q$. However, for $\alpha=2$
we get $\Energy(P,Q) = 2\| \E(X) - \E(Y) \|^2$, thus
$\Energy(P,Q)=0$ does 
not imply equality of distributions but only equality
of the means.

Consider data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ 
sampled from $k$ unknown distributions $\{ P_j \}_{j=1}^k$,
with points lying on a space of negative type,
$x_i \in \mathcal{X}$.
Let $\mathbb{X} = \bigcup_{j=1}^k \C_j$ be a disjoint
partition, $\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
can be empirically estimated with the aid of the
function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ is the number of elements in partition
$\C_i$. 
Define the \emph{within energy dispersion} as
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} as
\begin{equation}
\label{eq:between}
S \equiv \hspace{-1em}
\sum_{1 \le  i < j \le k } \hspace{-.5em} \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
A given point $x_i$ belongs to partition $\C_j$
if and only if $x_i \sim P_j$. 
The quantity $S$ is
a test statistic for equality of distributions
\citep{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_\ell \ne P_j$ for at least two $\ell\ne j$, 
we have that $S \to \infty$.

@@

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\citep{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \citep{Sejdinovic2013}.
When these conditions are satisfied we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
said to be equivalent kernels.
Now we can state the equivalence between energy distance and
inner products on RKHS, which is one of the main results of
\citep{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation}
\label{eq:Erho}
\begin{split}
\tfrac{1}{2}\Energy(P, Q) &= 
\E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y) \\ 
&= \gamma_\kk^2(P,Q) ,
\end{split}
\end{equation}
so we can compute the energy distance using the inner product
of $\Hk$.

@@

This section contains our main theoretical results, where 
we formulate an optimization problem for clustering 
based on energy statistics in the RKHS introduced in the previous section.
The proofs are contained in supplementary material.

@@

\begin{lemma}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ of negative type. 
For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes the between-sample statistic $S$, defined
in equation \eqref{eq:between}, if and only if
\begin{equation}
\label{eq:minimize}
\min_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k),
\end{equation}
where the within energy dispersion $W$ is defined by \eqref{eq:within}.
\end{lemma}

In the Euclidean case, the optimization problem \eqref{eq:minimize} based on
energy statistics was already proposed in \citep{Kgroups}. However, it is
important to note that this is equivalent to maximizing $S$,
which is the test statistic for equality of distributions. In this current
form, the relation with kernels and other clustering methods is obscure.
In the following, we show what is the explicit optimization problem behind 
\eqref{eq:minimize} in the corresponding RKHS, 
establishing the connection with kernel methods.

Assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$.  Define  the Gram matrix $G \in \mathbb{R}^{n\times n}$ as
\begin{equation}
\label{eq:kernel_matrix}
G_{ij} \equiv \kk(x_i,x_j).
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfies
$Z^\top Z = D$, where  
$D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. We also introduce the rescaled
matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} \
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases}
\end{equation}
Throughout the paper, we use the notation $M_{i\bullet}$ to denote
the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
Our next result shows that the optimization problem \eqref{eq:minimize}
is NP-hard since
it is a quadratically constrained quadratic program (QCQP) in the RKHS.

@@

Therefore, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
This optimization problem is nonconvex, and also NP-hard,
thus a direct approach 
is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints.
For instance, the relaxed problem
$\max_{Y} \Tr \left( Y^\top G \, Y \right)$ s.t. $Y^\top Y = I$,
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U \in \mathbb{R}^{n\times k}$ 
contain the top $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
\emph{Spectral clustering} is based on (variants of) this approach 
and will be compared to the iterative method that will be proposed 
in the following.

Note that the energy clustering problem \eqref{eq:qcqp2} 
is valid for data living in an \emph{arbitrary} space of negative type, where
a semimetric $\rho$, and thus the kernel $\kk$, are
assumed to be known. Standard energy statistics in
Euclidean spaces fixes a family of choices through 
\eqref{eq:rho_standard}.
The same would be valid for data living in more general
spaces $(\mathcal{X}, \rho)$.
In any case, energy  clustering 
is model-free, 
contrary to $k$-means and GMM, for example.
In practice, however,
the clustering quality strongly depend on the choice of a suitable
$\rho$ which measures the similarity between data points.
If prior information is available to choose $\rho$
it can be conveniently incorporated in the 
optimization problem \eqref{eq:qcqp2}.

\paragraph{Relation to Kernel $\bm{k}$-Means.}
One may wonder how energy clustering 
relates to the well-known kernel $k$-means problem%
\footnote{When we refer to kernel $k$-means problem we mean specifically 
the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
confused with kernel $k$-means algorithm that is just one possible recipe 
to solve \eqref{eq:kernel_kmeans}.}, 
extensively used in machine learning.
For a positive semidefinite Gram matrix $G$, as defined in
\eqref{eq:kernel_matrix},
there exists a feature map
$\varphi: \mathcal{X} \to \HH_\kk$ such that
$G_{ij} = \langle \varphi(x_i), \varphi(x_j) \rangle$. The kernel $k$-means optimization
problem,
in feature space,
is defined by
\begin{equation}
\label{eq:kernel_kmeans}
\min_{\C_1,\dotsc,\C_k}\bigg\{ 
\sum_{j=1}^k
\sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
\bigg\}
\end{equation}
where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
$\C_j$ in the ambient space $\mathcal{X}$. 
Notice that the above objective function
is strongly tied to the idea of minimizing distances between points
and cluster centers, which arises from $k$-means objective function based
on Lloyd's heuristic \citep{Lloyd}.
It is known \citep{Dhillon2,Dhillon}
that kernel $k$-means problem 
can be cast into a trace maximization in the same form as 
\eqref{eq:qcqp2}. The next result makes this explicit.

\begin{proposition}
\label{th:kernel_kmeans}
For a fixed kernel,
the energy clustering optimization problem
\eqref{eq:minimize} 
is equivalent to the kernel $k$-means optimization problem
\eqref{eq:kernel_kmeans}, and both are equivalent to \eqref{eq:qcqp2}.
\end{proposition}

@@

We introduce an iterative algorithm based
on Hartigan's method \citep{Hartigan} to find a local
maximizer of the optimization problem \eqref{eq:qcqp2}. Due to 
Proposition~\ref{th:kernel_kmeans} we can use 
kernel $k$-means algorithm \citep{Dhillon2,Dhillon}, which
will be compared with the proposed algorithm.

>>>>

\documentclass[twoside]{article}
\usepackage{aistats2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
%\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

@@

\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }

\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 }

%\aistatsauthor{Guilherme Fran\c{c}a \And  Joshua T. Vogelstein}

%\aistatsaddress{ Johns Hopkins University }

]

\begin{abstract}
Energy statistics introduces the notion of potential
energy between probability distributions, in  
close analogy to Newton's gravitational
potential in physics. We propose a principled approach to the
\mbox{clustering} problem based on energy statistics theory.
Our mathematical formulation establishes connection to kernel methods,
leading to a quadratically constrained quadratic program.
To obtain local solutions of such an optimization problem, which is NP-hard, 
we introduce an iterative
algorithm based
on Hartigan's method, which has the same computational cost but offers
several advantages compared to the 
well-known kernel $k$-means, based on Lloyd's heuristic.
We provide carefully designed numerical experiments showing
that our method is more flexible and outperforms
kernel $k$-means, spectral clustering,
standard $k$-means and Gaussian mixture models in a variety of settings,
specially in high \mbox{dimensions}. 
\end{abstract}

@@

Energy statistics \citep{Szkely2013}
provides a hypothesis test for equality of 
distributions, which is achieved 
under the minimum of the statistical potential energy. 
When probability distributions are different the 
statistical potential energy diverges as sample size increases, while tends 
to a nondegenerate limit distribution when probability
distributions are equal. 
The test statistic has compact representation
in terms of expectations of pairwise distances, providing
straightforward empirical estimates. Energy statistics
has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \citep{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \citep{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, an application of 
energy statistics to clustering in Euclidean spaces was 
considered \citep{Kgroups}.  
We refer to \citep{Szkely2013} for an overview.

@@

Although clustering from energy statistics in Euclidean spaces was considered
in \citep{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. Our results
immediately establish the connection between energy statistics based
clustering and kernel methods. Moreover, we show that that this 
QCQP is equivalent to kernel $k$-means optimization problem.

Our main 
algorithmic contribution is to use Hartigan's method \citep{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used in \citep{Kgroups}, but without any
connection to kernels. More importantly, the 
advantages of Hartigan's
over Lloyd's method was already demonstrated 
in some simple settings
\citep{Telgarsky,Slonin}, but apparently this method did not receive 
the deserved attention. To the 
best of our knowledge, Hartigan's method was not previously 
employed together with kernel methods. 
We provide a fully kernel based Hartigan's algorithm for clustering,
where the kernel is fixed by energy statistics. 
We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to solve our QCQP. We show that both algorithms  have the same
time complexity, however, Hartigan's method in kernel spaces is superior. 
Furthermore, in the examples considered in this paper, it 
also provides equal or better performance than spectral clustering,
which in fact solves a relaxed version of our QCQP.
Our numerical results provide compelling evidence that 
Hartigan's method applied to energy statistics based clustering
is more accurate and robust than 
kernel $k$-means. Moreover, we illustrate the
flexibility of energy clustering,  
showing that it is able to perform accurately on data coming from 
very different distributions, contrary to $k$-means and GMM for instance.
More specifically, the proposed method performs 
closely to $k$-means and GMM on normally distributed data, however,
it is significantly better on other settings. 
Its superiority in high dimensions is striking, being more accurate 
than $k$-means and GMM even on Gaussian settings.

@@

Consider random vectors $X_i$ living in an arbitrary space
$\mathcal{X}$ of \emph{negative type}, which means that $\mathcal{X}$
is endowed with a \emph{semimetric} 
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$ satisfying
$\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0$,
where $c_i \in \mathbb{R}$ and
$\sum_{i=1}^n c_i = 0$. 
Let $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments, and 
$X,X',Y,Y' \in \mathcal{X}$. 
The \emph{generalized energy distance} between $P$ and $Q$ is
given by 
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation} 
This quantity is nonnegative, $\Energy(P,Q) \ge 0$, and 
$\Energy^{1/2}$ is
a metric on the space of distributions. Energy distance provides
a characterization of equality of distributions.
For instance, the \emph{standard
energy distance} \citep{Szkely2013} in Euclidean spaces uses
the semimetric
\begin{equation}
\label{eq:rho_standard}
\rho_\alpha(X,Y) = \| X - Y\|^\alpha
\end{equation} 
where $0< \alpha \le 2$ and $\| \cdot \|$ is the
Euclidean norm in $\mathcal{X}=\mathbb{R}^D$. In this case $\Energy(P,Q)$
is rotationally invariant. For $0<\alpha<2$ we have
$\Energy(P,Q) = 0$ if and only if $P=Q$. However, for $\alpha=2$
we get $\Energy(P,Q) = 2\| \E(X) - \E(Y) \|^2$, thus
$\Energy(P,Q)=0$ does 
not imply equality of distributions but only equality
of the means.

Consider a sample $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ 
from $k$ unknown distributions $\{ P_j \}_{j=1}^k$,
with points lying on a space of negative type,
$x_i \in \mathcal{X}$.
Let $\mathbb{X} = \bigcup_{j=1}^k \C_j$ be a disjoint
partition, i.e. $\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
can be empirically estimated with the aid of the
function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ denotes the number of elements in partition
$\C_i$. 
Define the \emph{within energy dispersion} as
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} as
\begin{equation}
\label{eq:between}
S \equiv \hspace{-1em}
\sum_{1 \le  i < j \le k } \hspace{-.5em} \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
A given point $x_i$ belongs to partition $\C_j$
if and only if $x_i \sim P_j$. 
The quantity $S$ is
a test statistic for equality of distributions
\citep{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_\ell \ne P_j$ for at least two $\ell\ne j$, 
we have that $S \to \infty$.

@@

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\citep{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
and $x_0 \in \mathcal{X}$ be an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \citep{Sejdinovic2013}.
When these conditions are met we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
said to be equivalent kernels.
Now we can state the equivalence between energy distance and
inner products on RKHS, which is one of the main results of
\citep{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation}
\label{eq:Erho}
\begin{split}
\tfrac{1}{2}\Energy(P, Q) &= 
\E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y) \\ 
&= \gamma_\kk^2(P,Q) .
\end{split}
\end{equation}
Therefore, we can compute the energy distance using the inner product
of $\Hk$.

@@

This section contains our main theoretical results where 
we formulate an optimization problem for clustering 
based on energy statistics in the RKHS.
The proofs are contained in supplementary material.

@@

\begin{lemma}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ of negative type. 
For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes the between-sample statistic $S$, defined
in \eqref{eq:between}, if and only if
\begin{equation}
\label{eq:minimize}
\min_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k),
\end{equation}
where the within energy dispersion $W$ is defined by \eqref{eq:within}.
\end{lemma}

In the Euclidean case, the optimization problem \eqref{eq:minimize} based on
energy statistics was already proposed in \citep{Kgroups}. However, it is
important to note that this is equivalent to maximizing $S$,
which is the test statistic for equality of distributions. In this current
form, the relation with kernels and other clustering methods is obscure.
In the following we demonstrate what is the explicit optimization problem behind 
\eqref{eq:minimize} in the corresponding RKHS, 
establishing the connection with kernel methods.

Assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$.  Define  the Gram matrix $G \in \mathbb{R}^{n\times n}$ as
\begin{equation}
\label{eq:kernel_matrix}
G_{ij} \equiv \kk(x_i,x_j).
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfies
$Z^\top Z = D$, where  
$D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. We also introduce the rescaled
matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} \
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases}
\end{equation}
%Throughout the paper, we use the notation $M_{i\bullet}$ to denote
%the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
Our next result shows that the optimization problem \eqref{eq:minimize}
is NP-hard since
it is a quadratically constrained quadratic program (QCQP) in the 
associated RKHS.

@@

Therefore, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
This optimization problem is nonconvex, and NP-hard,
thus a direct approach 
is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints.
For instance, the relaxed problem
$\max_{Y} \Tr \left( Y^\top G \, Y \right)$ s.t. $Y^\top Y = I$,
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U \in \mathbb{R}^{n\times k}$ 
contain the top $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
\emph{Spectral clustering} is based on (variants of) this approach 
and will be compared to the iterative method that will be proposed 
in the next section.

Note that the energy clustering optimization problem \eqref{eq:qcqp2} 
is valid for data living in an \emph{arbitrary} space of negative type, where
a semimetric $\rho$, and thus the kernel $\kk$, are
assumed to be known. Standard energy statistics in
Euclidean spaces fixes a family of choices through 
\eqref{eq:rho_standard}.
The same would be valid for data living in more general
spaces $(\mathcal{X}, \rho)$.
In any case, energy  clustering 
is model-free, 
contrary to $k$-means and GMM, for example.
In practice, however,
the clustering quality strongly depends on the choice of a suitable
$\rho$ which measures the similarity between data points.
If prior information is available to make this choice
it can be immediately incorporated in the 
optimization problem \eqref{eq:qcqp2} through the kernel matrix $G$.

\paragraph{Relation to Kernel $\bm{k}$-Means.}
One may wonder how energy clustering 
relates to the well-known kernel $k$-means problem%
\footnote{When we refer to kernel $k$-means problem we mean specifically 
the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
confused with kernel $k$-means algorithm that is just one possible recipe 
to solve \eqref{eq:kernel_kmeans}. The distinction should be clear from 
the context.}, 
extensively used in machine learning.
For a positive semidefinite Gram matrix $G$, as defined in
\eqref{eq:kernel_matrix},
there exists a feature map
$\varphi: \mathcal{X} \to \HH_\kk$ such that
$G_{ij} = \langle \varphi(x_i), \varphi(x_j) \rangle$. 
The kernel $k$-means optimization
problem,
in feature space,
is defined by
\begin{equation}
\label{eq:kernel_kmeans}
\min_{\C_1,\dotsc,\C_k}\bigg\{ 
\sum_{j=1}^k
\sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
\bigg\}
\end{equation}
where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
$\C_j$ in the ambient space $\mathcal{X}$. 
Notice that the above objective function
is strongly tied to the idea of minimizing distances between points
and cluster centers, which arises from $k$-means objective function based
on Lloyd's heuristic \citep{Lloyd}.
It is known \citep{Dhillon2,Dhillon}
that kernel $k$-means problem 
can be cast into a trace maximization in the same form as 
\eqref{eq:qcqp2}. The next result makes this explicit.

\begin{proposition}
\label{th:kernel_kmeans}
For a fixed kernel,
the energy clustering optimization problem
\eqref{eq:minimize} 
is equivalent to kernel $k$-means optimization problem
\eqref{eq:kernel_kmeans}, and both are equivalent to \eqref{eq:qcqp2}.
\end{proposition}

@@

We now introduce an iterative algorithm based
on Hartigan's method \citep{Hartigan} to find a local
maximizer of the optimization problem \eqref{eq:qcqp2}. Due to 
Proposition~\ref{th:kernel_kmeans} we can also use 
kernel $k$-means algorithm \citep{Dhillon2,Dhillon} to solve
this optimization problem, and it
will be compared with the proposed algorithm in our numerical experiments.

We can rewrite the optimization problem \eqref{eq:qcqp2} as
\begin{equation}
\label{eq:maxQ}
\max_{\{ \C_1,\dotsc,\C_k \}} 
\bigg\{ Q = \sum_{j=1}^k \dfrac{Q_j}{n_j}  \bigg\},
\quad Q_j \equiv \sum_{x,y\in\C_j} \kk(x,y),
\end{equation}
where $Q_j$ represents an internal energy cost of cluster $\C_j$, and
$Q$ is the total energy cost where each $Q_j$ 
is weighted by the inverse
of the number of points in $\C_j$. For a data point $x_i$ we denote
its own energy cost
with the entire cluster $\C_\ell$ by
\begin{equation}
\label{eq:costxij}
Q_\ell(x_i) \equiv \sum_{y\in\C_\ell} \kk(x_i, y) = 
G_{i \bullet} \cdot Z_{\bullet \ell},
\end{equation}
where $G_{i\bullet}$ ($G_{\bullet i}$) denotes
the $i$th row (column) of the matrix $G$.
